The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-12-04 categorical-zero-one.tex
add lemma, fix minor error in proof of causality

    It follows that $\prod_i X_i$ is an infinite tensor product in $\Kl(H)$; the preservation condition is easily verified since adding another factor to this product again gives the infinite product. The Kolmogorov product condition is similarly immediate to verify.
\end{proof}
\begin{proposition}
    $\Kl(H)$ is causal.
\end{proposition}
\begin{proof}
    We are given spaces $A, X, Y, Z$, and continuous maps $f:A \to HX, g: X \to HY, h_1,h_2: Y \to HZ$.
    We assume that two maps $X \to H(W \times Z)$ are equal.
    They are given by
    \[a \mapsto \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_1(y) \times \{y\}\right),\ x \mapsto \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_2(y) \times \{y\}\right)\]
    Of course, equality of these maps simply means that the right-hand sides agree for each $x$.

@@

    The desired conclusion is that the sets
    \[\cl\left(\bigcup_{x \in f(a), y \in g(x)} h_1(y) \times \{y\} \times \{x\}\right),\ \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_2(y) \times \{y\} \times \{x\}\right)\]
    agree. Let us rewrite them as
    \[\cl\left(\{(z,y,x) \mid z \in h_1(y), y\in g(x), x\in f(a)\}\right),\ \cl\left(\{(z,y,x) \mid z \in h_2(y), y\in g(x), x\in f(a)\}\right)\]

@@

    Let us take a point $(z_0,y_0,x_0)$ left-hand set, and consider an open neighborhood $U_1 \times U_2 \times U_3$, for open subsets of $Z,Y,X$ respectively.
    Our goal is to construct $(z,y,x) \in U_1\times U_2 \times U_3$ so that $x \in f(a), y \in g(x), z \in h_2(y)$.

    The assumptions on $(z_0,y_0,x_0)$ implies that the point $(z_0,y_0)$ is in $S$.
    Hence by the other expression of $S$, we can find a point $x$ so that $x \in f(a), y_0 \in g(x) \cap U_2, z_0 \in h_2(y_0) \cap U_1$.
    The issue is now to fix this so that $x \in f(a) \cap U_3$.

    Consider the set $\{y \mid h_2(y) \cap U_1 \neq \emptyset\}$.
    This is an open set which contains our $y_0$ above. Let $U_2'$ be the (nonempty!) intesection of this set with $U_2$.
    Now consider $\{x \mid g(x) \cap U_2' \neq \emptyset\}$. This is again an open set which cantains $x_0$. Let $U_3'$ be the intersection of this with $U_3$.
    Now since $x_0 \in f(a)$, any open neighborhood of $x_0$ intersects $f(a)$ (since it's closed).
    So choose $x_1 \in f(a) \cap U_3', y_1 \in U_2' \cap g(x) \cap U_2',$ and finally $z_1 \in h_2(y) \cap U_1$.
    (Each possible by the construction of the previous element). By construction we have $(z_1,y_1,x_1) \in U_1 \times U_2 \times U_3$, exactly as desired.
    The other inclusion follows by a symmetric argument.
\end{proof}

>>>>

    It follows that $\prod_i X_i$ is an infinite tensor product in $\Kl(H)$; the preservation condition is easily verified since adding another factor to this product again gives the infinite product. The Kolmogorov product condition is similarly immediate to verify.
\end{proof}
\begin{proposition}\label{prop:klh-causal}
    $\Kl(H)$ is causal.
\end{proposition}
The proof of this proposition is fairly technical and not very illuminating. We recommend skipping to \Cref{vietoris_kolmogorov} below on a first reading.
We will use the notation of \cite{hyperspace} through the proof.

First, we will need a quick lemma.
\begin{lemma}\label{lem:extraclosure}
    Let $f: X \to HY$ and $g: Y \to HZ$ be continuous functions
    Consider the map $HX \to HZ$ given by composing $f$ and $g$ in $\Kl(H)$, yielding a map $X \to HZ$, then lifting this using the monad structure.
    By definition, this map sends a closed $A \subset X$ to
    \[\cl \left(\bigcup_{a \in A} \cl\left(\bigcup_{b \in f(a)} g(b)\right)\right)\]

    In fact, this equals \[\cl\left(\bigcup_{a \in A}\bigcup_{b \in f(a)} g(b)\right)\]
\end{lemma}
The point of this lemma is that the iterated closures that arise when composing maps in $\Kl(H)$ can be mostly ignored, although we do have to take a closure at the end.
\begin{proof}
    Let $\hat{f},\hat{g}$ denote the induced maps $HX \to HY, HY \to HZ$.
    Since both of the sets we are comparing are closed, it suffices to show that they intersect the same collection of open sets.
    We use the notation $\braket{U,V}$ from \cite{hyperspace}, which is defined for two subsets of the same space to be $1$ if they intersect and $0$ otherwise.\footnote{we generalize slightly and use this notation even when $V$ is not closed, although $U$ is still always open.}

    Now we calculate, for a given open $U \subset Z$
    \[\braket{U , \hat{g}(\hat{f}(A))} = \braket{U , \cl(\cup_{y \in \hat{f}(A)}g(y))} = \braket{ U, \cup_{y \in \hat{f}(A)}g(y)}\]
    Here we first insert the definition of $\hat{g}$, then use the fact that $U$ is open.

    Now the statement that this union intersects $U$ means just that at least one of the sets in the union does.
    So we get
    \[=\braket{g\inv(\Hit(U)) , \hat{f}(A)}\]
    Note that $g\inv(\Hit(U))$ is still an open set, by continuity.
    Applying these arguments again, we find
    \[=\braket{f\inv(\Hit(g\inv(\Hit(U)))) , A} = \braket{\Hit(g\inv(\Hit(U))) , f(A)}\]
    where $f(A) \subset HY$ is simply the direct image of $A$.
    The statement that $f(A)$ intersects $\Hit(g\inv(\Hit(U)))$ means that for at least one $a\in A$, $f(a)$ is in $\Hit(g\inv(\Hit(U)))$, i.e it intersects $g\inv(\Hit(U))$.
    So we can pass to
    \[= \braket{g\inv(\Hit(U)) , \cup_{x \in A} f(x)}\]
    And by the same argument, to
    \[= \braket{U , \cup_{x \in A}\cup_{y\in f(x)}g(x)} = \braket{U , \cl \cup_{x \in A}\cup_{y\in f(x)}g(x)}\]
    which is exactly what we wanted.
\end{proof}

\begin{proof}[Proof of \Cref{prop:klh-causal}]
    We are given spaces $A, X, Y, Z$, and continuous maps $f:A \to HX, g: X \to HY, h_1,h_2: Y \to HZ$.
    We assume that two maps $X \to H(W \times Z)$ are equal.
    They are given by
    \[a \mapsto \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_1(y) \times \{y\}\right),\ x \mapsto \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_2(y) \times \{y\}\right)\]
    Of course, equality of these maps simply means that the right-hand sides agree for each $x$.
    Here we have implictly used \Cref{lem:extraclosure} to get rid of extra closures.

@@

    The desired conclusion is that the sets
    \[\cl\left(\bigcup_{x \in f(a), y \in g(x)} h_1(y) \times \{y\} \times \{x\}\right),\ \cl\left(\bigcup_{x \in f(a), y \in g(x)} h_2(y) \times \{y\} \times \{x\}\right)\]
    agree. Let us rewrite them as
    \[\cl\left(\{(z,y,x) \mid z \in h_1(y), y\in g(x), x\in f(a)\}\right),\ \cl\left(\{(z,y,x) \mid z \in h_2(y), y\in g(x), x\in f(a)\}\right)\]
    (Where we have again used \Cref{lem:extraclosure}).

@@

    Let us take a point $(z_0,y_0,x_0)$ in the left-hand set, and consider an open neighborhood $U_1 \times U_2 \times U_3$, for open subsets of $Z,Y,X$ respectively.
    Our goal is to construct $(z,y,x) \in U_1\times U_2 \times U_3$ so that $x \in f(a), y \in g(x), z \in h_2(y)$.

    The assumptions on $(z_0,y_0,x_0)$ implies that the point $(z_0,y_0)$ is in $S$.
    Hence by the other expression of $S$, we can find points $x_1,y_1,z_1$ so that $x_1 \in f(a), y_1 \in g(x) \cap U_2, z_1 \in h_2(y_1) \cap U_1$.
    The issue is now to fix this so that $x \in f(a) \cap U_3$.

    Consider the set $\{y \mid h_2(y) \cap U_1 \neq \emptyset\}$.
    This is an open set which contains our $y_1$ above. Let $U_2'$ be the (nonempty!) intesection of this set with $U_2$.
    Now consider $\{x \mid g(x) \cap U_2' \neq \emptyset\}$. This is again an open set which contains $x_1$. Let $U_3'$ be the intersection of this with $U_3$.
    Now since $x_1 \in f(a)$, any open neighborhood of $x_1$ intersects $f(a)$ (since it's closed).
    So choose $x_2 \in f(a) \cap U_3', y_2 \in U_2' \cap g(x_2) \cap U_2',$ and finally $z_2 \in h_2(y_2) \cap U_1$.
    (Each possible by the construction of the previous element). By construction we have $(z_2,y_2,x_2) \in U_1 \times U_2 \times U_3$, exactly as desired.
    The other inclusion follows by a symmetric argument.
\end{proof}
